---
title: "Interactive Tree"
author: "Niel Infante"
output: html_document
---

This document will take you through creating an interactive tree plot of your microbiome data.



```{r setup, include=F}
knitr::opts_chunk$set(echo = T, message=F, warning=F)

```
### Load the needed packages.

```{r packages}
library(phyloseq)
library(tidyverse)
library(plotly)
library(data.table)
library(scales)

```

### Read the needed data, turn it into a phyloseq object

Some data munging needs to happen to get the data files we have into the form needed for phyloseq. Ignore these steps if your data is well formed, or if you already have a phyloseq object

```{r read_data}

# Read Data
otu <- read_csv('Data_Files/otu.csv')
taxa <- read_csv('Data_Files/taxa.csv')
meta <- read_csv('Data_Files/Tara_oceans_mapping_file.csv')
tree <- ape::read.tree('Data_Files/tree.tre')

# Get OTU ready
otu <- as.data.frame(otu)
row.names(otu) <- otu$X1
otu$X1 <- NULL
otu <- otu_table(otu, taxa_are_rows = T)
  
# Get Taxonomy ready
taxonomy <- as.data.frame(taxa)
row.names(taxonomy) <- taxonomy$X1
taxonomy$X1 <- NULL
taxonomy <- as.matrix(taxonomy)
taxonomy <- tax_table(taxonomy)

# Get metadata ready
meta <- as.data.frame(meta)
row.names(meta) <- meta$SampleID
meta <- sample_data(meta)

# Create Phyloseq Object
ps <- phyloseq(otu, taxonomy, meta, tree)

```

### Define tree function

This is the plot tree function from phyloseq. However, I made a few small changes to make it play nicely with plotly. You will not need to change this for your data. Run this step, but you don't have to read all this code. I did so you don't have to, just like Paul McMurdie and Susan Holmes originally wrote the code so I didn't have to. My change was to add tooltip as a passed parameter.

```{r define_func}

interactive_plot_tree <- function (physeq, method = "sampledodge", nodelabf = NULL, color = NULL, 
          shape = NULL, size = NULL, min.abundance = Inf, label.tips = NULL, 
          text.size = NULL, sizebase = 5, base.spacing = 0.02, ladderize = FALSE, 
          plot.margin = 0.2, title = NULL, treetheme = NULL, justify = "jagged", tooltip = NULL) 
{
  fix_reserved_vars = function(aesvar) {
    aesvar <- gsub("^abundance[s]{0,}$", "Abundance", aesvar, 
                   ignore.case = TRUE)
    aesvar <- gsub("^OTU[s]{0,}$", "OTU", aesvar, ignore.case = TRUE)
    aesvar <- gsub("^taxa_name[s]{0,}$", "OTU", aesvar, ignore.case = TRUE)
    aesvar <- gsub("^sample[s]{0,}$", "Sample", aesvar, ignore.case = TRUE)
    return(aesvar)
  }
  if (!is.null(label.tips)) {
    label.tips <- fix_reserved_vars(label.tips)
  }
  if (!is.null(color)) {
    color <- fix_reserved_vars(color)
  }
  if (!is.null(shape)) {
    shape <- fix_reserved_vars(shape)
  }
  if (!is.null(size)) {
    size <- fix_reserved_vars(size)
  }
  if (is.null(phy_tree(physeq, FALSE))) {
    stop("There is no phylogenetic tree in the object you have provided.\n", 
         "Try phy_tree(physeq) to see for yourself.")
  }
  if (!inherits(physeq, "phyloseq")) {
    method <- "treeonly"
  }
  treeSegs <- tree_layout(phy_tree(physeq), ladderize = ladderize)
  edgeMap = aes(x = xleft, xend = xright, y = y, yend = y)
  vertMap = aes(x = x, xend = x, y = vmin, yend = vmax)
  p = ggplot(data = treeSegs$edgeDT) + geom_segment(edgeMap) + 
    geom_segment(vertMap, data = treeSegs$vertDT)
  if (is.null(text.size)) {
    text.size <- phyloseq:::manytextsize(ntaxa(physeq))
  }
  if (!is.null(label.tips) & method != "sampledodge") {
    labelDT = treeSegs$edgeDT[!is.na(OTU), ]
    if (!is.null(tax_table(object = physeq, errorIfNULL = FALSE))) {
      taxDT = data.table(tax_table(physeq), OTU = taxa_names(physeq), 
                         key = "OTU")
      labelDT = merge(x = labelDT, y = taxDT, by = "OTU")
    }
    if (justify == "jagged") {
      labelMap <- aes_string(x = "xright", y = "y", label = label.tips, 
                             color = color)
    }
    else {
      labelMap <- aes_string(x = "max(xright, na.rm=TRUE)", 
                             y = "y", label = label.tips, color = color)
    }
    p <- p + geom_text(labelMap, data = labelDT, size = I(text.size), 
                       hjust = -0.1, na.rm = TRUE)
  }
  if (is.null(nodelabf)) {
    nodelabf = phyloseq:::howtolabnodes(physeq)
  }
  p = nodelabf(p, treeSegs$edgeDT[!is.na(label), ])
  p = nodelabf(p, treeSegs$vertDT[!is.na(label), ])
  if (is.null(treetheme)) {
    treetheme <- theme(axis.ticks = element_blank(), axis.title.x = element_blank(), 
                       axis.text.x = element_blank(), axis.title.y = element_blank(), 
                       axis.text.y = element_blank(), panel.background = element_blank(), 
                       panel.grid.minor = element_blank(), panel.grid.major = element_blank())
  }
  if (inherits(treetheme, "theme")) {
    p <- p + treetheme
  }
  if (!is.null(title)) {
    p <- p + ggtitle(title)
  }
  if (method != "sampledodge") {
    return(p)
  }
  dodgeDT = treeSegs$edgeDT[!is.na(OTU), ]
  dodgeDT = merge(x = dodgeDT, y = data.table(psmelt(physeq), 
                                              key = "OTU"), by = "OTU")
  if (justify == "jagged") {
    dodgeDT <- dodgeDT[Abundance > 0, ]
  }
  if (!is.null(color) | !is.null(shape) | !is.null(size)) {
    setkeyv(dodgeDT, cols = c("OTU", color, shape, size))
  }
  else {
    setkey(dodgeDT, OTU, Sample)
  }
  dodgeDT[, `:=`(h.adj.index, 1:length(xright)), by = OTU]
  if (justify == "jagged") {
    dodgeDT[, `:=`(xdodge, (xright + h.adj.index * base.spacing * 
                              max(xright, na.rm = TRUE)))]
  }
  else {
    dodgeDT[, `:=`(xdodge, max(xright, na.rm = TRUE) + h.adj.index * 
                     base.spacing * max(xright, na.rm = TRUE))]
    dodgeDT <- dodgeDT[Abundance > 0, ]
  }
  dodgeMap <- aes_string(x = "xdodge", y = "y", color = color, 
                         fill = color, shape = shape, size = size, names=tooltip)
  p <- p + geom_point(dodgeMap, data = dodgeDT, na.rm = TRUE)
  if (!is.null(size)) {
    p <- p + scale_size_continuous(trans = log_trans(sizebase))
  }
  if (any(dodgeDT$Abundance >= min.abundance[1])) {
    pointlabdf = dodgeDT[Abundance >= min.abundance[1], ]
    p <- p + geom_text(mapping = aes(xdodge, y, label = Abundance), 
                       data = pointlabdf, size = text.size, na.rm = TRUE)
  }
  if (!is.null(label.tips)) {
    tiplabDT = dodgeDT
    tiplabDT[, `:=`(xfartiplab, max(xdodge)), by = OTU]
    tiplabDT <- tiplabDT[h.adj.index == 1, .SD, by = OTU]
    if (!is.null(color)) {
      if (color %in% sample_variables(physeq, errorIfNULL = FALSE)) {
        color <- NULL
      }
    }
    labelMap <- NULL
    if (justify == "jagged") {
      labelMap <- aes_string(x = "xfartiplab", y = "y", 
                             label = label.tips, color = color)
    }
    else {
      labelMap <- aes_string(x = "max(xfartiplab, na.rm=TRUE)", 
                             y = "y", label = label.tips, color = color)
    }
    p <- p + geom_text(labelMap, tiplabDT, size = I(text.size), 
                       hjust = -0.1, na.rm = TRUE)
  }
  min.x <- -0.01
  max.x <- dodgeDT[, max(xright, na.rm = TRUE)]
  if ("xdodge" %in% names(dodgeDT)) {
    max.x <- dodgeDT[, max(xright, xdodge, na.rm = TRUE)]
  }
  if (plot.margin > 0) {
    max.x <- max.x * (1 + plot.margin)
  }
  p <- p + scale_x_continuous(limits = c(min.x, max.x))

  return(p)
}
```

### Subset data

I reduce the data just so the plot is smaller and prettier. Subset or not as you want to explore your data. subset_taxa also is helpful.

```{r subset}

list_to_keep <- as.logical( c(rep(c(1,0,0,0,0), 27), 0, 1, 0, 0))

ps_small <- prune_samples(list_to_keep, ps)

#ps_small <- subset_taxa(ps, Order=='Oceanospirillales')

```

### Make reporting taxa

This is a holdover from when I was working with more incomplete data. I leave it here because it does no harm, and illustrated some of what you might do. Going forward I will be using Genus. 
Here I change the taxa table so that there is a column called lowest_taxa. This is the taxonomic classification of the leaf with the lowest known name. NA's are removed, as are entries such as f__. 
 
```{r report}

# Collapse into single column
# This data has no Species column. If yours does, the index should be 1:7
Lowest_taxa <- do.call(paste, c(as.data.frame(tax_table(ps_small))[1:6], sep=";"))

Lowest_taxa <- gsub("(;NA)*$", "", Lowest_taxa, perl=T)    # Remove NAs
Lowest_taxa <- gsub("(;.__)*$", "", Lowest_taxa, perl=T)   # Remove g__
Lowest_taxa <- gsub("^.*;", "", Lowest_taxa, perl=T)       # Remove everything to the left of the remaining entry

# Add back to tax_table
colnames(tax_table(ps_small))[1] <- 'Lowest_taxa'
tax_table(ps_small)[,'Lowest_taxa'] <- Lowest_taxa

```

This is not necessary, you can use "Family", or whatever taxonomic level you like, by specifying that everywhere you see "Lowest_taxa" below.


### Draw Tree and make interactive

```{r draw_tree}

pt <- interactive_plot_tree(ps_small, ladderize="left", color="Depth", size='abundance', nodelabf=nodeplotblank, base.spacing=0.1, tooltip = 'Genus')
  
plotly_tree<-ggplotly(p=pt, tooltip = 'Genus') 

plotly_tree

```

This draws the tree. The dots are the different samples we have, colored by sample type. The size of the dot is a measure of abundance. Unfortunately, plotly overwrites color choices, and removes the abundance scale. I will update if I figure out how to fix this.

More information on how to tweek the tree graph can be found [here](https://joey711.github.io/phyloseq/plot_tree-examples.html)

You can save as an html using the below.


```{r save, eval=F}
htmlwidgets::saveWidget(plotly_tree, 'tree_with_hover.html')
```



